LCMM, PLS-DA, and LDA exploration

### 2016-04-10 doc/manuscript.Rmd
Added DC title page + added much to discussion (old edits)

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl:
    - /home/luke/Documents/bibliography/styles/vancouver.csl
output:
    word_document:
      reference_docx: /home/luke/Templates/journal-templates/ajcn-template.docx
      fig_height: 6 
      fig_width: 5
      fig_caption: true
---

# Title page #

Title: Total non-esterified fatty acids, but not composition, predict
progression in beta-cell dysfunction

Author: Luke W. Johnston, Richard P. Bazinet, Anthony J. Hanley

Affiliations:

Disclaimers:

Correspondence:

Funding support:

# Abstract #

@@

Overall story: while nefa are known risk factors for diabetes, not sure about
the composition. we find that the composition doesn't matter so much as the
overall pool, specific to BCD. however, 16:0, etc, are the only two fa that have
an assoc, suggesting that irrespective of other fa increases or decreases, these
two fa are the fa that contribute to the harm seen in the overall nefa. using
cluster analysis, we see that those with low bcd have a pattern of fa that skews
to higher amounts of 16:0 etc and marginally less in other fa (maybe not enough
to observe on an individual scale). while other studies have seen a role of nefa
in ir, we do not. while this warrants further longitudinal research, this may
suggest that given prev studies on ir have been acute in nature, over the long
term the role of nefa on ir may not be substantial. this duality in nefa is also
seen in bcd, where acutely it improves bc, but chronically no.

- as GEE results suggest that % does nothing, but in conc 16:0 etc has assoc...
is it that 16:0 is harmful or is it the pool? is it that as 16:0 inc, there is
harm, irrespective of whether other FA increase or decrease as well?

figs/tabs: 
- tab:basic char
- fig:distr
- fig:nefaCorr (sup?)
- fig:varCorr (sup?)
- fig:geeconc%
- fig:plsda

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?

@@

```{r knit_setup, collapse=TRUE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = FALSE)
```

```{r setup, message=FALSE}
source('.Rprofile')
suppressMessages(run_setup())
load_data()
ds <- ds %>% 
    mutate(lALT = log(ALT),
           lTAG = log(TAG))
dim(ds)
```

```{r caption_setup}
## Include captions below using `captioner` package
fig <- captioner(prefix = 'Figure')
cite_f <- pryr::partial(fig, display = 'cite')
sfig <- captioner(prefix = 'Supplementary Figure')
cite_sf <- pryr::partial(sfig, display = 'cite')
tab <- captioner(prefix = 'Table')
cite_t <- pryr::partial(tab, display = 'cite')
stab <- captioner(prefix = 'Supplementary Table')
cite_st <- pryr::partial(stab, display = 'cite')
# usage: cite_st('basicChar')

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 time points.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006)')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for {{covariates}}. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values presented were adjusted for the false discovery rate.')

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that
exposure of cell or animal to high concentrations of NEFA can induce insulin
resistance in insulin-sensitive tissue such as the muscle or liver, impair
beta-cell production of insulin, and cause hyperglycemia {{confirm...cite}}.
Epidemiological and clinical studies have observed similar findings, showing
that elevated total plasma NEFA associates with risk for diabetes {{cite}}.

@@

```{r variableSet}
outcomes_is <- c('linvHOMA', 'lISI')
outcomes_bcf <- c('lIGIIR', 'lISSI2')
outcomes <- c(outcomes_is, outcomes_bcf)
ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'MET', 'AlcoholPerWk')
covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')

gee_df <- get_gee_data(ds)
```

```{r geeData, cache=TRUE}
gee_results <- bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
    )

gee_results_unadj <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
    )
```

@@

```{r inlineResults}
pctne <- calculate_percent_nefa_contribution(ds) 
misclass_npct <- calculate_plsda_misclass(plsda_results)
pctChg_o <- calculate_outcomes_pct_change(ds)
lcmm_grps <- calculate_ngroups_lcmm(lcmm_results)
```

@@

The primary outcome variables declined between `r pctChg_o` over the 6 years.

@@

Several associations were seen in the unadjusted GEE models 
(`r cite_sf('gee')`), particularly for the concentration data. Full model
adjustment attenuated most of these associations, with total NEFA, 16:0, and
18:2n-6 (all as nmol/mL) had negative associations between IGI/IR and ISSI-2
(see `r cite_f('gee')`). The magnitude of association for each of these
variables was fairly consistent for each beta-cell function measure. For every
one SD decrease in any of these three NEFA variables, there was a predicted
{{magnitude}}% lower IGI/IR and a {{magnitude}}% lower ISSI-2 at each clinic
visit. None of the NEFA variables as a mol% had any association with the
outcomes. Adjusting for TG attenuates all associations with IGI/IR and ISSI-2
(data not shown), though sample size at the 6 year visit is restricted to n=128
in those who had TG measured. There were no significant interaction effects
between time and the individual NEFA on any of the outcome measures (data not
shown). Raw values from the GEE models are shown in `r cite_st('gee_unadj')` for
unadjusted models and `r cite_st('gee')` for fully adjusted models.

@@

Three latent classes were extracted from LCMM, with `r lcmm_grps[2,2]` 
participants in the high, `r lcmm_grps[1,2]` in the middle, and `lcmm_grps[3,2]`
in the low beta-cell function groups. The trajectories of the three groups are
shown in `r cite_sf('lcmm')`. These groups were used as the response variables
in the PLS-DA analysis. The PLS-DA results show a very poor discriminatory
abililty of the NEFA composition in classifying participants into the correct
beta-cell function latent class. Only `r misclass_npct[2,1]` 
(`r misclass_npct[2,4]`) participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function.
`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
extracted component, while `r cite_sf('plsda_groups')` shows the clustering of the components by ISSI-2 latent class.

@@

In a Canadian population of adults who are at-risk for diabetes, we found that
higher total NEFA independently predicted a lower beta-cell function over 6
years. In particular, the specific composition of the NEFA fraction had no
association with any of the metabolic outcomes within this cohort. These
findings suggest that the size of the total fraction, rather than the
composition, influences the pathogenesis of diabetes.

Given we found associations with the nmol/mL data, but none in the mol% data,
this suggests to us that the total size of the NEFA fraction is more important
in the pathogenesis of diabetes than is the specific FA composition. A
limitation of the mol% data is the inherent dependency between the FA as an
increase in one FA by definition means other FA must decrease. In the context of
this dependency, the associations we found for 16:0 and 18:2n-6 as nmol/mL
suggests that these NEFA may be contributing to the harmful association seen for
the total NEFA fraction, irrespective of changes in the other NEFA.

## Beta cell

We found that higher palmitic acid (16:0), oleic acid (18:1n-9), and linoleic
acid (18:2n-6) associated with lower beta-cell function by {{number?}}. Expand
on experimental studies, pathology/biology, mechanisms, etc.
[@Salgin2012a;@Giacca2011a;@Xiao2009a]

In a cross-sectional analysis of the RISCK cohort, total NEFA had a negative
association with insulin sensitivity and, particularly strong with beta-cell
function [@Johns2014a]. The authors further analysis showed that greater
adiposity was not a mediator between the total NEFA and metabolic function
association.

Experimental [@Carpentier2003a]
Review [@Carpentier2000a;@Cnop2008]

Most experimental studies on beta-cell function use only palmitic acid....

First study to our knowledge to examine a broad spectrum of individual NEFA in a cohort setting on beta-cell function..

We found no association of any individual NEFA as mol% on beta-cell function,
but a very strong signal with total NEFA, palmitic acid, oleic acid, and
linoleic acid as concentrations.

Using cluster techniques, we found no evidence to suggest that any individual
NEFA contributed uniquely to beta-cell function, confirming the results we saw
in the GEE models. This is particularly seen with the poor predictive ability of
the NEFA composition to classify between groups of beta-cell function.

@@

## IR

There is substantial experimental evidence highlighting the role of increased 
NEFA and a subsequent increase in insulin resistance via impairment of insulin 
signaling cascades (reviewed in multiple articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a;@Surampudi2009]). However, in this
longitudinal analysis, we saw no association between any individual or total
NEFA and neither hepatic (1/HOMA-IR) nor whole-body insulin sensitivity (ISI). 
There are some possible explanations for our findings. Previous studies done *in
vivo* were mostly short term infusion 
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
[@Daniele2014a;@Liang2013a] trials and the observational research was generally
cross-sectional [@Miller2012a;@Johns2014a;@Bush2012a]. The limitation of these
studies were the short time periods examined and the experimentally-induced
elevations in NEFA concentrations, making it difficult to assess the causal role
in non-experimental environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a; @Johns2014a], suggesting that either
NEFA only have a minor role in IR or that NEFA is the effect, but not cause, of
the underlying IR. Adipose tissue, which release NEFA during fasting, can become
insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA update [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA is a consequence of higher IR.

Another explanation for our null findings with insulin sensitivity may be
differences in physiology between fasting and postprandial NEFA kinetics. For
instance, some experimental studies using the clamp found that fasting NEFA was
a weak predictor of insulin sensitivity compared to postprandial concentrations
of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the
adipose tissue following postprandial TAG lipolysis via lipoprotein lipase
may result in NEFA spillover into the blood and a subsequent increase in circulating
NEFA [@Almandoz2013a], which may be more metabolically active given postprandial
activity.

...Last possibility is that given these subjects were at-risk for diabetes already,
the causal role of NEFA on IR may have long passed and it no longer influences
IR....

## Composition

This is the first study to our knowledge to examine the composition of the NEFA
fraction and longitudinal measures of metabolic functioning.

Composition of NEFA and metabolic syndrome components [@Dai2015a]

Composition of NEFA and metabolic syndrome binary [@Novgorodtseva2011a]

Using complex measurement and statistical techniques that were rigorously
applied {{? reword}} to the composition of the NEFA fraction, Liu *et al.*
-[@Liu2010a] found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful biomarkers
for identifying healthy compared to diabetic individuals. However, a limitation
of this study was the smaller sample size {{confirm}} and the cross-sectional
design.

Limitations, etc...

Final, brief conclusion...

## Acknowledgements ##

# Tables #

```{r tabBasic}
# table(ds$Ethnicity, ds$f.VN)
# table(ds$f.VN)
summary(ds$Sex)
ds %>%
    outline_table(c('ne_Total', 'HOMA', 'ISI', 'IGIIR', 'ISSI2',
                    'TAG', 'Chol', 'BMI', 'Waist', 'HDL', 'FamHistDiab',
                    'Age', 'Ethnicity', 'Sex', 'MET', 'IFG', 'IGT',
                    'DM'), 'f.VN') %>% 
#     add_rows(c('HOMA', 'ISI', 'IGIIR', 'ISSI2'), stat_medianIQR, digits = 1) %>% 
#     add_rows(c('TAG', 'Chol', 'HDL', 'ne_Total', 'BMI', 'Waist', 'Age'), 
#              stat_meanSD, digits = 1) %>% 
    add_rows(c('Ethnicity', 'Sex')) %>% 
    rename_rows(renaming_table_rows) %>%
    rename_columns('Measure', 'Baseline', '3-yr', '6-yr') %>%
    construct_table()

```

# Figures #

@@

```{r supFigLDA}

```

@@

```{r supFigLCMM, fig.cap=}
plot_lcmm_results(lcmm_data)
```

```{r supFigPlsLoadings, fig.cap=}
plot_plsda_loadings(plsda_results)
```

```{r supFigPlsGrp}
plot_plsda_grouping(plsda_results)
```

>>>>

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
csl:
    - /home/luke/Documents/bibliography/styles/vancouver.csl
output:
    md_document:
        variant: markdown
    word_document:
        reference_docx: diabetes-care-template.docx
        fig_height: 6 
        fig_width: 5
        fig_caption: true
---

# Title page

Title: Total non-esterified fatty acids, but not composition, predict
progression in beta-cell dysfunction: The Prospective Metabolism and Islet Cell
Evaluation (PROMISE) cohort

<!-- 47 characters + spaces -->
Running title: NEFA and the pathogenesis of diabetes

<!-- First, middle initial, last, highest academic degree -->
Author: Luke W. Johnston, Stewart Harris, Ravi Retnakaran, Bernard Zinman,
Zhen Liu, Richard P. Bazinet, Anthony J. Hanley

<!-- During time of study. -->
Affiliation:

Corresponding author:

- Name: Anthony J. Hanley
- Current address:
- Phone number:
- Fax number:
- Email: anthony.hanley@utoronto.ca

Word count: / 4000

Table and figure count: / 4

<!-- Reference limit / 40 -->

<!-- 250 words -->
# Abstract

Objective:

Research Design and Methods:

Results:

Conclusions:

# Abstract

@@

TODO:
- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
the p-value...?
- What about adiponectin?
- Precise followup time for each participant?
- Interaction by sex or ethnicity

@@

```{r knit_setup, echo=FALSE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(echo = FALSE, dev = c('png', 'postscript'))
```

```{r setup, message=FALSE}
source('.Rprofile')
suppressMessages(run_setup())
ds <- load_data()
```

```{r caption_setup}
## Include captions below using `captioner` package
fig <- captioner(prefix = 'Figure')
cite_f <- pryr::partial(fig, display = 'cite')
sfig <- captioner(prefix = 'Supplemental Figure S')
cite_sf <- pryr::partial(sfig, display = 'cite')
tab <- captioner(prefix = 'Table')
cite_t <- pryr::partial(tab, display = 'cite')
stab <- captioner(prefix = 'Supplemental Table S')
cite_st <- pryr::partial(stab, display = 'cite')
# usage: cite_st('basicChar')

tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006)')
fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.  Adjusted for {{covariates}}. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values presented were adjusted for the false discovery rate.')

@@

Total non-esterified fatty acids (NEFA) have been well documented to influence
the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that
exposure of cells or animals to high concentrations of NEFA can induce insulin
resistance in insulin-sensitive tissue such as the muscle or liver and impair
beta-cell production of insulin, and cause hyperglycemia [@Rachek2014a;@Cnop2008a].
Epidemiological and clinical studies have observed similar findings, showing
that elevated total plasma NEFA associates with risk for diabetes [@Daniele2014a {{more}}].

@@

```{r variableSet}
gee_df <- get_gee_data(ds)

outcomes_is <- c('linvHOMA', 'lISI')
outcomes_bcf <- c('lIGIIR', 'lISSI2')
outcomes <- c(outcomes_is, outcomes_bcf)
ne_pct <- grep('^pct_ne', names(gee_df), value = TRUE)
ne_conc <- c('TotalNE', grep('^ne\\d\\d', names(gee_df), value = TRUE))
covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'MET', 'AlcoholPerWk')
covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')
```

```{r geeData, cache=TRUE}
gee_results <- dplyr::bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
    ) %>% 
    tidy_gee_results()

gee_results_unadj <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
    analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
    ) %>% 
    tidy_gee_results()
```

@@

```{r inlineResults}
percent.nefa <- calculate_percent_nefa_contribution(ds) 
misclass.n.percent <- calculate_plsda_misclass(plsda_results)
percent.change.outcomes <- calculate_outcomes_pct_change(ds)
lcmm.groups <- calculate_ngroups_lcmm(lcmm_results)
```

@@

The primary outcome variables had a median decline between 
`r percent.change.outcomes` over the 6 years in this subset of the PROMISE cohort.

@@

Several associations were seen in the unadjusted GEE models 
(`r cite_sf('gee')`), particularly for the concentration data. Full model
adjustment attenuated most of these associations, where total NEFA, 16:0, and
18:2n-6 (all as nmol/mL) had negative associations between IGI/IR and ISSI-2
(see `r cite_f('gee')`). The magnitude of association for each of these
variables was fairly consistent for each beta-cell function measure. For every
one SD decrease in any of these three NEFA variables, there was a predicted
{{magnitude}}% lower IGI/IR and a {{magnitude}}% lower ISSI-2 at each clinic
visit. None of the NEFA variables as a mol% had any association with the
outcomes. Adjusting for TG attenuates all associations with IGI/IR and ISSI-2
(data not shown), though sample size at the 6 year visit is restricted to n=128
in those who had TG measured. There were no significant interaction effects
between time and the individual NEFA on any of the outcome measures (data not
shown). Raw values from the GEE models are shown in `r cite_st('gee_unadj')` for
unadjusted models and `r cite_st('gee')` for fully adjusted models.

@@

Three latent classes were extracted from LCMM, with `r lcmm_grps[2,2]` 
participants in the high, `r lcmm_grps[1,2]` in the middle, and `lcmm_grps[3,2]`
in the low beta-cell function groups. The trajectories of the three groups are
shown in `r cite_sf('lcmm')`. These groups were used as the response variables
in the PLS-DA analysis. The PLS-DA results show a very poor discriminatory
abililty of the NEFA composition in classifying participants into the correct
beta-cell function latent class. Only `r misclass_npct[2,1]` 
(`r misclass_npct[2,4]`) participants were correctly classified, primarily into the
middle group. The specific NEFA composition was not able to accurately classify
participants who had low or high beta-cell function.
`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
extracted component and `r cite_sf('plsda_groups')` shows the clustering of the
components by ISSI-2 latent class.

@@

In a Canadian population of adults who are at-risk for diabetes, we found that
higher total NEFA independently predicted a lower beta-cell function over 6
years. In particular, the specific composition of the NEFA fraction had no
association with any of the metabolic outcomes within this cohort. These
findings suggest that the size of the total fraction, rather than the
composition, influences the pathogenesis of diabetes.
Given we found associations with the nmol/mL data, but none in the mol% data,
this suggests to us that the total size of the NEFA fraction is more important
in the pathogenesis of diabetes than is the specific FA composition. A
limitation of the mol% data is the inherent dependency between the FA as an
increase in one FA by definition means other FA must decrease. In the context of
this dependency, the associations we found for 16:0 and 18:2n-6 as nmol/mL
suggests that these NEFA may be contributing to the harmful association seen for
the total NEFA fraction, irrespective of changes in the other NEFA.

While future studies are needed to confirm these findings, our results suggest
that measurement of the NEFA composition may not be warranted and that the
simple, easy to measure total NEFA may be the measure most appropriate for
further analysis. {{reword}}

The role of total NEFA on diabetes is a well-studied area of research.
Epidemiology study showing that higher NEFA in both adults and children 
associate with lower insulin secretion [@Salgin2012a]. Also show an association 
with higher risk for developing IGT or diabetes after approximately 6 years in 
adults [@Salgin2012a].
In a cross-sectional analysis of the RISCK cohort, total NEFA had a negative
association with insulin sensitivity and, particularly strong with beta-cell
function [@Johns2014a]. The authors' further analysis suggested that greater
adiposity was likely not a mediator between the total NEFA and metabolic
function association.
{{[@Giacca2011a;@Xiao2009a]}}.
This is the first study to our knowledge to examine a broad spectrum of
individual NEFA in a cohort setting on beta-cell function. We found that
individual NEFA as mol% or of the NEFA composition overall had any association
or predictive power on determining beta-cell function, but there was a very
strong signal with total NEFA, palmitic acid, oleic acid, and linoleic acid as
concentrations.
Potential mechanisms for the role of NEFA on beta-cell function are well
described experimentally, particularly for palmitic acid. Prolonged exposure to
the elevated NEFA can induce apoptosis in the beta-cells, likely through 
endoplasmic reticulum stress, formation of ceramides, and generation of nitric 
oxide, as well as can impair proinsulin production and mitochrondial function 
[@Cnop2008a; @Maris2013a; @Carpentier2003a].

@@

There is substantial experimental evidence highlighting the role of increased 
NEFA and a subsequent increase in insulin resistance via impairment of insulin 
signaling cascades (reviewed in multiple articles 
[@Martins2012a;@Ebbert2013a;@Capurso2012a;@Surampudi2009]). However, in this
longitudinal analysis, we saw no association between any individual or total
NEFA and neither hepatic (1/HOMA-IR) nor whole-body insulin sensitivity (ISI). 
There are some possible explanations for our findings. Previous studies done *in
vivo* were mostly short term infusion 
[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
[@Daniele2014a;@Liang2013a] trials and the observational research was generally
cross-sectional [@Miller2012a;@Johns2014a;@Bush2012a]. The limitation of these
studies were the short time periods examined and the experimentally-induced
elevations in NEFA concentrations, making it difficult to assess the causal role
in non-experimental environments over long time frames. Moreover, some studies
have found null or weak associations of NEFA and IR 
[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a;@Johns2014a], suggesting that either
NEFA only have a minor role in IR or that NEFA is the effect, but not cause, of
the underlying IR. Adipose tissue, which release NEFA during fasting, can become
insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
lipase may not respond to the inhibitory action of insulin. As such, IR may 
develop before elevations in blood NEFA. There is also evidence to suggest that 
insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
the hypothesis that elevated NEFA is a consequence of higher IR.

There are two other possible explanations for our null findings for the insulin
sensitivity measures. First, there may be differences in physiology between
fasting and postprandial NEFA kinetics. For instance, some experimental studies
using the clamp found that fasting NEFA was a weak predictor of insulin
sensitivity compared to postprandial concentrations of NEFA
[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
tissue following postprandial TAG lipolysis via lipoprotein lipase may result in
NEFA spillover into the blood and a subsequent increase in circulating NEFA
[@Almandoz2013a], which may be more metabolically active given postprandial 
activity. Second, the null findings may be due to the high risk population
examined. It may be that in this population, IR has become well established and
the role of NEFA may no longer be evaluated in this context.

This is the first study to our knowledge to examine the composition of the NEFA
fraction and longitudinal measures of metabolic functioning. A recent, well
analyzed study used a variety of advanced statistical techniques to explore the
multivariate relationship between the NEFA composition and components of the
metabolic syndrome (MetS) [@Dai2015a]. Specifically, using the statistical techniques,
the authors identified the NEFA 16:1n-9, 20:1n-9, and 22:4n-6 to most correlate
with components of the MetS. Another statistically similar studied diabetes
specifically and found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful
biomarkers for identifying healthy compared to diabetic individuals [@Liu2010a].
However, a limitation of both these studies were the smaller sample sizes
(approximately 100 subjects) and the cross-sectional design.

There are a few important limitations to our study. NEFA were only quantified at
the baseline visit and as such we can not investigate whether there were 
concomitant changes in NEFA and the metabolic measures. However, we believe this
is a strength for our specific objective, as the chance of reverse causality is 
reduced given that fatty acid and glucose metabolism pathways are tightly 
integrated. This is also an observational cohort, and there may be some residual
confounding we haven't considered or that couldn't be measured. Nonetheless, we 
were as thorough as possible in understanding and minimizing potential 
confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
who are primarily female of European-ancestry and as such our results may not be
generalizable to other populations. However, given these limitations, our study
also has several strengths, including the longitudinal design and the rigorous
statistical techniques and methods applied, which are specifically suited to
investigating temporal relationships and to handling the multivariate nature of
the data. In addition, our cohort contains highly detailed and comprehensive
variable measurements at each collection visit, and has both concentration and
mol% data for the fatty acids.

Final, brief conclusion...
In conclusion, we found that NEFA
While future studies are needed to confirm these findings, our results suggest
that measurement of the NEFA composition may not be warranted and that the
simple, easy to measure total NEFA may be the measure most appropriate for
further analysis. {{reword}}

## Acknowledgements

The authors thank Jan Neuman, Paula Van Nostrand, Stella Kink, and Annette
Barnie of the Leadership Sinai Centre for Diabetes, Mount Sinai Hospital,
Toronto, Canada and Sheila Porter and Mauricio Marin of the Centre for Studies
in Family Medicine, University of Western Ontario, London, Canada for their 
expert technical assistance and dedication in their work for PROMISE. 
The authors had the following responsibility: LWJ conducted research, analyzed
data, and wrote the paper; RR, ZL, BZ, and SBH designed research, conducted
research, and provided essential materials (infrastructure and clinical
resources); RR, BZ, SBH, RPB, and AG provided intellectual feedback on the
paper; RPB conducted research, provided essential reagents and materials; AJH
designed research, assisted with interpretation, and provided intellectual
feedback on all versions of the paper; LWJ and AJH had primary responsibility
for final content. All authors read and approved the final manuscript.  The
authors report no potential conflicts of interest relevant to this study.

# Tables

```{r tabBasic}
table_basic(ds, caption = tab_basic)
```

Note: Values are in median (IQR), mean (SD), and n (%).

# Figures

@@

@@

```{r supFigLCMM, fig.cap=sfig_lcmm}
plot_lcmm_results(lcmm_data)
```

```{r supFigPlsLoadings, fig.cap=sfig_plsda_loadings}
plot_plsda_loadings(plsda_results)
```

```{r supFigPlsGrp, fig.cap=sfig_plsda_groups}
plot_plsda_grouping(plsda_results)
```
